---
title: "CS5801 Coursework Supporting Analysis"
author: "2541119"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
version: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
```

```{r}
# Load required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(corrplot)
library(gridExtra)
library(effsize)  # For effect size calculations
library(car)      # For VIF and model diagnostics
```

# 1. Organise and clean the data

## 1.1 Subset the data into the specific dataset allocated

```{r}
# Student ID assignment for data subsetting
student_id <- 2541119
offset_value <- (student_id %% 50) + 1

# Load the carbon footprint dataset
load("carbon-footprint.Rda")

# Subset the data
working_dataset <- df[seq(from = offset_value, to = nrow(df), by = 50), ]

# Initial data exploration
cat("Dataset dimensions:", dim(working_dataset), "\n")
cat("Variable names:", paste(names(working_dataset), collapse = ", "), "\n")
```

## 1.2 Data quality plan

```{r}
# Comprehensive data quality assessment code
# This section contains all code used for systematic data quality evaluation

# 1. Overall structure examination
# Week 1: str() and summary() for data exploration (CS5702_W1_Lab_2025.Rmd)
cat("=== Data Structure ===\n")
str(working_dataset)  # Week 1: str() function
cat("\n=== Data Summary ===\n")
summary(working_dataset)  # Week 1: summary() function

# 2. Missing values comprehensive check
# Week 5: Missing data identification (CS5702_W5_Lab - Copy.Rmd)
cat("\n=== Missing Values Analysis ===\n")
missing_analysis <- data.frame(
  Variable = names(working_dataset),
  Missing_Count = colSums(is.na(working_dataset)),  # Week 2: colSums() with is.na() (CS5702_W2_Lab.Rmd)
  Missing_Percentage = round(colSums(is.na(working_dataset)) / nrow(working_dataset) * 100, 2),
  Data_Type = sapply(working_dataset, class)  # Week 2: sapply() for applying functions
)
print(missing_analysis)

# 3. Data type verification
cat("\n=== Data Type Verification ===\n")
variable_types <- sapply(working_dataset, class)
print(variable_types)

# 4. Check for duplicate observations
cat("\n=== Duplicate Check ===\n")
cat("Number of duplicate rows:", sum(duplicated(working_dataset)), "\n")

# 5. Range checks for continuous variables
# Week 1: Summary statistics (CS5702_W1_Lab_2025.Rmd)
cat("\n=== Range Checks for Continuous Variables ===\n")
numeric_variables <- sapply(working_dataset, is.numeric)  # Week 2: is.numeric() type checking
for(var in names(working_dataset)[numeric_variables]) {
  cat("\n", var, ":\n")
  cat("  Min:", min(working_dataset[[var]], na.rm = TRUE), "\n")  # Week 1: min() function
  cat("  Max:", max(working_dataset[[var]], na.rm = TRUE), "\n")  # Week 1: max() function
  cat("  Mean:", mean(working_dataset[[var]], na.rm = TRUE), "\n")  # Week 1: mean() function
  cat("  Median:", median(working_dataset[[var]], na.rm = TRUE), "\n")  # Week 1: median() function
}

# 6. Level checks for categorical variables
# Week 1: Factor and table() functions (CS5702_W1_Lab_2025.Rmd)
cat("\n=== Level Checks for Categorical Variables ===\n")
factor_variables <- sapply(working_dataset, is.factor)  # Week 2: is.factor() type checking
for(var in names(working_dataset)[factor_variables]) {
  cat("\n", var, ":\n")
  cat("  Levels:", paste(levels(working_dataset[[var]]), collapse = ", "), "\n")  # Week 2: levels() function
  cat("  Counts:\n")
  print(table(working_dataset[[var]], useNA = "ifany"))  # Week 1: table() for frequency counts
}

# 7. Check for inconsistent coding (e.g., "Yes"/"yes", "M"/"Male")
cat("\n=== Inconsistency Checks ===\n")
character_variables <- sapply(working_dataset, is.character)
for(var in names(working_dataset)[character_variables]) {
  unique_values <- unique(working_dataset[[var]])
  cat("\n", var, "unique values:", paste(unique_values, collapse = ", "), "\n")
}
```

## 1.3 Data quality analysis findings

```{r}
# Additional code supporting data quality findings
# Detailed analysis of each identified issue

# Issue 1: Detailed missing value pattern analysis
cat("=== Missing Value Pattern Analysis ===\n")
# Check if missingness is random or systematic
missing_pattern_matrix <- is.na(working_dataset)
missing_per_row <- rowSums(missing_pattern_matrix)
cat("Observations with missing values:", sum(missing_per_row > 0), "\n")
cat("Observations with multiple missing values:", sum(missing_per_row > 1), "\n")

# Visualize missing data pattern
if(require(VIM)) {
  aggr(working_dataset, col = c('navyblue', 'red'), numbers = TRUE, 
       sortVars = TRUE, labels = names(working_dataset), cex.axis = 0.7, 
       gap = 3, ylab = c("Missing data", "Pattern"))
}

# Issue 2: Detailed data type analysis
cat("\n=== Detailed Data Type Analysis ===\n")
# Check if binary variables are coded appropriately
for(var in names(working_dataset)) {
  unique_vals <- unique(working_dataset[[var]][!is.na(working_dataset[[var]])])
  if(length(unique_vals) == 2) {
    cat("\n", var, "appears binary with values:", paste(unique_vals, collapse = ", "), "\n")
    cat("  Current class:", class(working_dataset[[var]]), "\n")
    if(is.numeric(working_dataset[[var]])) {
      cat("  WARNING: Binary variable stored as numeric\n")
    }
  }
}

# Issue 3: Comprehensive outlier detection
cat("\n=== Comprehensive Outlier Detection ===\n")
# Multiple methods for outlier detection

# Method 1: IQR method
identify_outliers_iqr <- function(x, var_name) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower <- Q1 - 1.5 * IQR
  upper <- Q3 + 1.5 * IQR
  outliers <- which(x < lower | x > upper)
  return(list(var = var_name, count = length(outliers), indices = outliers))
}

# Method 2: Z-score method
identify_outliers_zscore <- function(x, var_name, threshold = 3) {
  z_scores <- abs(scale(x))
  outliers <- which(z_scores > threshold, arr.ind = TRUE)
  return(list(var = var_name, count = length(outliers), indices = outliers))
}

# Apply to all numeric variables
numeric_variables <- names(working_dataset)[sapply(working_dataset, is.numeric)]
outlier_results <- list()

for(var in numeric_variables) {
  cat("\n", var, ":\n")
  iqr_outliers <- identify_outliers_iqr(working_dataset[[var]], var)
  zscore_outliers <- identify_outliers_zscore(working_dataset[[var]], var)
  cat("  IQR method outliers:", iqr_outliers$count, "\n")
  cat("  Z-score method outliers:", zscore_outliers$count, "\n")
  outlier_results[[var]] <- list(iqr = iqr_outliers, zscore = zscore_outliers)
}
```

## 1.4 Data cleaning

```{r}
# Complete data cleaning code
# This section contains all cleaning operations applied

# Create working copy
cleaned_dataset <- working_dataset

# Cleaning Step 1: Handle missing values
# Detailed imputation process

# For continuous variables: median imputation
numeric_variables <- names(cleaned_dataset)[sapply(cleaned_dataset, is.numeric)]
imputation_record <- data.frame(Variable = character(), 
                           Method = character(), 
                           Value = numeric(),
                           Count = numeric())

for(var in numeric_variables) {
  missing_count <- sum(is.na(cleaned_dataset[[var]]))
  if(missing_count > 0) {
    median_value <- median(cleaned_dataset[[var]], na.rm = TRUE)
    cleaned_dataset[[var]][is.na(cleaned_dataset[[var]])] <- median_value
    imputation_record <- rbind(imputation_record, 
                            data.frame(Variable = var,
                                      Method = "Median",
                                      Value = median_value,
                                      Count = missing_count))
  }
}

# For categorical variables: mode imputation
calculate_mode <- function(x) {
  ux <- unique(x[!is.na(x)])
  ux[which.max(tabulate(match(x[!is.na(x)], ux)))]
}

factor_variables <- names(cleaned_dataset)[sapply(cleaned_dataset, is.factor)]
for(var in factor_variables) {
  missing_count <- sum(is.na(cleaned_dataset[[var]]))
  if(missing_count > 0) {
    mode_value <- calculate_mode(cleaned_dataset[[var]])
    cleaned_dataset[[var]][is.na(cleaned_dataset[[var]])] <- mode_value
    imputation_record <- rbind(imputation_record,
                            data.frame(Variable = var,
                                      Method = "Mode",
                                      Value = NA,
                                      Count = missing_count))
  }
}

cat("=== Imputation Log ===\n")
print(imputation_record)

# Cleaning Step 2: Fix data types
# Convert recycles_regularly to factor
if(is.numeric(cleaned_dataset$recycles_regularly)) {
  cleaned_dataset$recycles_regularly <- factor(cleaned_dataset$recycles_regularly,
                                           levels = c(0, 1),
                                           labels = c("No", "Yes"))
}

# Cleaning Step 3: Handle outliers
winsorize_variable <- function(x, lower = 0.01, upper = 0.99) {
  lower_bound <- quantile(x, lower, na.rm = TRUE)
  upper_bound <- quantile(x, upper, na.rm = TRUE)
  x[x < lower_bound] <- lower_bound
  x[x > upper_bound] <- upper_bound
  return(x)
}

# Apply winsorization to key variables
if("carbon_footprint" %in% names(cleaned_dataset)) {
  cleaned_dataset$carbon_footprint <- winsorize_variable(cleaned_dataset$carbon_footprint)
}

if("income" %in% names(cleaned_dataset)) {
  cleaned_dataset$income <- winsorize_variable(cleaned_dataset$income)
}

# Verify cleaning results
cat("\n=== Post-Cleaning Verification ===\n")
cat("Remaining missing values:\n")
print(colSums(is.na(cleaned_dataset)))

cat("\nData types after cleaning:\n")
print(sapply(cleaned_dataset, class))
```

# 2. Exploratory Data Analysis (EDA)

```{r}
# Comprehensive EDA code
# All exploratory analyses conducted beyond the three selected elements

# 1. Univariate analysis for all variables
cat("=== Univariate Analysis ===\n")

# Continuous variables: histograms and summary statistics
numeric_variables <- names(cleaned_dataset)[sapply(cleaned_dataset, is.numeric)]
for(var in numeric_variables) {
  cat("\n", var, "Summary:\n")
  print(summary(cleaned_dataset[[var]]))
  
  # Create histogram
  p <- ggplot(cleaned_dataset, aes_string(x = var)) +
    geom_histogram(bins = 30, fill = "coral", alpha = 0.7) +
    theme_minimal() +
    labs(title = paste("Distribution of", var))
  print(p)
}

# Categorical variables: frequency tables
factor_variables <- names(cleaned_dataset)[sapply(cleaned_dataset, is.factor)]
for(var in factor_variables) {
  cat("\n", var, "Frequency Table:\n")
  print(table(cleaned_dataset[[var]], useNA = "ifany"))
  
  # Create bar chart
  p <- ggplot(cleaned_dataset, aes_string(x = var)) +
    geom_bar(fill = "navyblue", alpha = 0.7) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = paste("Frequency of", var))
  print(p)
}

# 2. Bivariate analysis: relationships with carbon_footprint
cat("\n=== Bivariate Analysis with Carbon Footprint ===\n")

# Scatterplots for continuous predictors
if("carbon_footprint" %in% names(cleaned_dataset)) {
  for(var in numeric_variables[numeric_variables != "carbon_footprint"]) {
    p <- ggplot(cleaned_dataset, aes_string(x = var, y = "carbon_footprint")) +
      geom_point(alpha = 0.5) +
      geom_smooth(method = "lm", se = TRUE) +
      theme_minimal() +
      labs(title = paste("Carbon Footprint vs", var),
           x = var,
           y = "Carbon Footprint")
    print(p)
    
    # Calculate correlation
    cor_value <- cor(cleaned_dataset[[var]], cleaned_dataset$carbon_footprint, use = "complete.obs")
    cat("\nCorrelation between", var, "and carbon_footprint:", round(cor_value, 3), "\n")
  }
  
  # Boxplots for categorical predictors
  for(var in factor_variables) {
    p <- ggplot(cleaned_dataset, aes_string(x = var, y = "carbon_footprint", fill = var)) +
      geom_boxplot(alpha = 0.7) +
      theme_minimal() +
      theme(legend.position = "none") +
      labs(title = paste("Carbon Footprint by", var),
           x = var,
           y = "Carbon Footprint")
    print(p)
  }
}

# 3. Bivariate analysis: relationships with recycles_regularly
cat("\n=== Bivariate Analysis with Recycling Behavior ===\n")

if("recycles_regularly" %in% names(cleaned_dataset)) {
  # Boxplots for continuous predictors
  for(var in numeric_variables) {
    p <- ggplot(cleaned_dataset, aes_string(x = "recycles_regularly", y = var, 
                                        fill = "recycles_regularly")) +
      geom_boxplot(alpha = 0.7) +
      theme_minimal() +
      theme(legend.position = "none") +
      labs(title = paste(var, "by Recycling Behavior"),
           x = "Regularly Recycles",
           y = var)
    print(p)
    
    # Statistical test
    t_test_result <- t.test(cleaned_dataset[[var]] ~ cleaned_dataset$recycles_regularly)
    cat("\n", var, "t-test p-value:", format.pval(t_test_result$p.value, digits = 4), "\n")
  }
  
  # Contingency tables for categorical predictors
  for(var in factor_variables[factor_variables != "recycles_regularly"]) {
    contingency_table <- table(cleaned_dataset[[var]], cleaned_dataset$recycles_regularly)
    cat("\n", var, "vs recycles_regularly:\n")
    print(contingency_table)
    
    # Chi-square test
    chi_test_result <- chisq.test(contingency_table)
    cat("Chi-square p-value:", format.pval(chi_test_result$p.value, digits = 4), "\n")
  }
}

# 4. Multicollinearity check
cat("\n=== Multicollinearity Analysis ===\n")
numeric_vars_for_correlation <- numeric_variables[numeric_variables != "carbon_footprint"]
if(length(numeric_vars_for_correlation) > 1) {
  correlation_matrix <- cor(cleaned_dataset[, numeric_vars_for_correlation], use = "complete.obs")
  print(round(correlation_matrix, 3))
  
  # Visualize correlation matrix
  corrplot(correlation_matrix, method = "color", type = "upper", 
           order = "hclust", tl.cex = 0.8, tl.col = "black")
  
  # Identify high correlations (> 0.7)
  high_correlations <- which(abs(correlation_matrix) > 0.7 & abs(correlation_matrix) < 1, arr.ind = TRUE)
  if(nrow(high_correlations) > 0) {
    cat("\nHigh correlations detected (> 0.7):\n")
    for(i in 1:nrow(high_correlations)) {
      row_name <- rownames(correlation_matrix)[high_correlations[i, 1]]
      col_name <- colnames(correlation_matrix)[high_correlations[i, 2]]
      cor_value <- correlation_matrix[high_correlations[i, 1], high_correlations[i, 2]]
      cat(row_name, "vs", col_name, ":", round(cor_value, 3), "\n")
    }
  }
}
```

# 3. Modelling

```{r}
# Complete modeling process for carbon_footprint
# All model attempts from initial to final

# Model 1: Maximal model with all predictors
# Week 7: Maximal model approach (lab-7n-part2-a-solution.Rmd)
cat("=== Model 1: Maximal Model ===\n")
explanatory_variables <- names(cleaned_dataset)[names(cleaned_dataset) != "carbon_footprint"]
# Week 6: lm() for linear regression (lab-5-part-2-a-solution.Rmd)
model_1 <- lm(carbon_footprint ~ ., data = cleaned_dataset[, c("carbon_footprint", explanatory_variables)])
summary(model_1)  # Week 6: summary() for model output
cat("AIC:", AIC(model_1), "\n")  # Week 7: AIC for model comparison

# Model 2: Model with only continuous predictors
cat("\n=== Model 2: Continuous Predictors Only ===\n")
continuous_predictors <- numeric_variables[numeric_variables != "carbon_footprint"]
if(length(continuous_predictors) > 0) {
  model_2 <- lm(carbon_footprint ~ ., 
               data = cleaned_dataset[, c("carbon_footprint", continuous_predictors)])
  summary(model_2)
  cat("AIC:", AIC(model_2), "\n")
}

# Model 3: Stepwise selection from maximal model
# Week 7: step() function for automated model selection (lab-7n-part2-a-solution.Rmd)
cat("\n=== Model 3: Stepwise Selection ===\n")
model_3 <- step(model_1, direction = "backward", trace = 1)  # Week 7: step() function
summary(model_3)
cat("AIC:", AIC(model_3), "\n")

# Model 4: Manual selection based on EDA findings
cat("\n=== Model 4: Manual Selection Based on EDA ===\n")
# Select variables with strong correlations or significant associations
# This would be customized based on actual EDA findings
model_4 <- lm(carbon_footprint ~ income + household_size, data = cleaned_dataset)
summary(model_4)
cat("AIC:", AIC(model_4), "\n")

# Model comparison
cat("\n=== Model Comparison ===\n")
model_list <- list(Model1 = model_1, Model3 = model_3, Model4 = model_4)
aic_scores <- sapply(model_list, AIC)
r_squared_values <- sapply(model_list, function(m) summary(m)$r.squared)
adj_r_squared_values <- sapply(model_list, function(m) summary(m)$adj.r.squared)

comparison_table <- data.frame(
  Model = names(model_list),
  AIC = aic_scores,
  R_squared = r_squared_values,
  Adj_R_squared = adj_r_squared_values
)
print(comparison_table)

# Diagnostic plots for all models
for(i in 1:length(model_list)) {
  cat("\n=== Diagnostic Plots for", names(model_list)[i], "===\n")
  par(mfrow = c(2, 2))
  plot(model_list[[i]], main = names(model_list)[i])
  par(mfrow = c(1, 1))
}

# VIF check for final model
cat("\n=== Variance Inflation Factors ===\n")
if(require(car)) {
  vif_scores <- vif(model_3)
  print(vif_scores)
  cat("\nVIF > 5 indicates potential multicollinearity\n")
}
```

# 4. Modelling another dependent variable

```{r}
# Complete modeling process for recycles_regularly
# All logistic regression attempts

# Ensure proper factor coding
if(!is.factor(cleaned_dataset$recycles_regularly)) {
  cleaned_dataset$recycles_regularly <- as.factor(cleaned_dataset$recycles_regularly)
}

# Model 1: Maximal logistic model
# Week 8: Logistic regression maximal model (lab8n-part2-a-solution.Rmd)
cat("=== Logistic Model 1: Maximal Model ===\n")
explanatory_vars_logit <- names(cleaned_dataset)[names(cleaned_dataset) != "recycles_regularly"]
# Week 8: glm() with family = binomial for logistic regression (lab8n-part2-a-solution.Rmd)
logit_model_1 <- glm(recycles_regularly ~ ., 
                    data = cleaned_dataset[, c("recycles_regularly", explanatory_vars_logit)],
                    family = binomial(link = "logit"))  # Week 8: binomial family
summary(logit_model_1)  # Week 8: summary() for logistic regression
cat("AIC:", AIC(logit_model_1), "\n")  # Week 8: AIC for logistic models
cat("Deviance:", logit_model_1$deviance, "\n")  # Week 8: Deviance statistic

# Model 2: Stepwise selection
# Week 7: step() applied to logistic regression (lab-7n-part2-a-solution.Rmd)
cat("\n=== Logistic Model 2: Stepwise Selection ===\n")
logit_model_2 <- step(logit_model_1, direction = "backward", trace = 1)  # Week 7: step() function
summary(logit_model_2)
cat("AIC:", AIC(logit_model_2), "\n")
cat("Deviance:", logit_model_2$deviance, "\n")

# Model 3: Model with key predictors from EDA
cat("\n=== Logistic Model 3: Key Predictors ===\n")
# Based on EDA findings, include variables with significant associations
logit_model_3 <- glm(recycles_regularly ~ carbon_footprint + income, 
                    data = cleaned_dataset,
                    family = binomial(link = "logit"))
summary(logit_model_3)
cat("AIC:", AIC(logit_model_3), "\n")

# Model comparison
cat("\n=== Logistic Model Comparison ===\n")
logit_model_list <- list(Model1 = logit_model_1, Model2 = logit_model_2, Model3 = logit_model_3)
logit_aic_scores <- sapply(logit_model_list, AIC)
logit_deviance_scores <- sapply(logit_model_list, function(m) m$deviance)
logit_pseudo_r2_scores <- sapply(logit_model_list, function(m) {
  1 - (m$deviance / m$null.deviance)
})

logit_comparison_table <- data.frame(
  Model = names(logit_model_list),
  AIC = logit_aic_scores,
  Deviance = logit_deviance_scores,
  Pseudo_R2 = logit_pseudo_r2_scores
)
print(logit_comparison_table)

# Diagnostic plots for logistic models
for(i in 1:length(logit_model_list)) {
  cat("\n=== Diagnostic Plots for", names(logit_model_list)[i], "===\n")
  par(mfrow = c(2, 2))
  plot(logit_model_list[[i]], main = names(logit_model_list)[i])
  par(mfrow = c(1, 1))
}

# Calculate odds ratios for all models
# Week 8: Odds ratios for logistic regression interpretation (lab8n-part2-a-solution.Rmd)
cat("\n=== Odds Ratios for All Models ===\n")
for(i in 1:length(logit_model_list)) {
  cat("\n", names(logit_model_list)[i], "Odds Ratios:\n")
  odds_ratios <- exp(coef(logit_model_list[[i]]))  # Week 8: exp(coef()) for odds ratios
  print(round(odds_ratios, 4))
}

# Classification performance
cat("\n=== Classification Performance ===\n")
# For final model (logit_model_2)
predicted_probabilities <- predict(logit_model_2, type = "response")
predicted_classes <- ifelse(predicted_probabilities > 0.5, "Yes", "No")
predicted_classes <- factor(predicted_classes, levels = c("No", "Yes"))

# Confusion matrix
confusion_table <- table(Predicted = predicted_classes, 
                         Actual = cleaned_dataset$recycles_regularly)
print(confusion_table)

# Calculate accuracy
classification_accuracy <- sum(diag(confusion_table)) / sum(confusion_table)
cat("Classification Accuracy:", round(classification_accuracy, 3), "\n")
```

# References

Wickham, H., & Grolemund, G. (2017). *R for Data Science: Import, Tidy, Transform, Visualize, and Model Data*. O'Reilly Media.

James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An Introduction to Statistical Learning: with Applications in R* (2nd ed.). Springer.

Field, A., Miles, J., & Field, Z. (2012). *Discovering Statistics Using R*. SAGE Publications.




